Machine learning topological phases in real space

### 2020-07-02 8_prb_submission/1_reply_referee/article_with_comments/article.tex
Finished section information entropy signatures in the macroscopic space

\section*{Uncertainty relations}
In this section we explore some theoretical possibilities that arise from the information entropy signatures presented in the previous section. The discussion presented here is complemented by the section \textbf{Information entropy signatures in the macroscopic limit} in the Supplementary Material.

@@

The interpretation of the information entropy signatures of the previous section in terms of information entropy wave functions opens several avenues for investigation. Here we explore its most forthright corollary, which is the establishment of uncertainty relations for topological phase transitions. The counterpart of the spatial information entropy wave function $\psi_S(x)$ in wavevector space is its Fourier transform 
\begin{equation}
\label{fourier_shannon_wave_function}
\hat{\psi}_S(k) = \int_{0}^{1}\psi_S(x)e^{-2\pi ikx}dx
\end{equation}
from which the Shannon information entropy density function can be computed in wavevector space,
\begin{equation}
\label{shannon_wave_function_wavevector_space}
\hat{\rho}_S(k) = |\hat{\psi}_S(k)|^2.
\end{equation}
In possession of this quantum formalism, we may write two canonical uncertainty relations binding together the real space and wavevector space information entropy density functions \ref{shannon_wave_function} and \ref{shannon_wave_function_wavevector_space} for topological phase transitions:
\begin{enumerate}
\item[i)] \emph{The Heisenberg uncertainty principle}
\begin{equation}
\Delta\rho_S \cdot \Delta\hat{\rho}_S \geq \frac{1}{4\pi} \quad
\end{equation}
and 
\item[ii)] \emph{The Hirschman entropic uncertainty}
\begin{equation}
H[\rho_S] + H[\hat{\rho}_S] \geq \log\Big(\frac{e}{2}\Big). \quad
\end{equation}
\end{enumerate}

@@

\textcolor{blue}{We should also compare our results to those obtained in \cite{ml_interpretability}, where the subject of investigation is the interpretability of neural network models trained to recognize topological phase transitions in some physical systems. In that paper, interesting visualizations are shown demonstrating that the patterns captured by a single-layer feedforward neural network indeed correspond to known physical quantities that are relevant to the problems at hand. This should indeed be called "model interpretability", as in that case the authors introspect into their models to make sure that they are learning patterns of physical pertinence to the systems at hand. In our paper we prefer the term "model explainability", as we are using similar model introspection tools to propose previously unknown concepts and properties of the physical systems being investigated. While the nuances in the meanings of these two terms are the subject of often heated philosophical debates in the artificial intelligence community, this choice of nomenclature suits the practical application of these model introspection techniques to physics well.}\citequote{superior_comparable}

@@

>>>>

\section*{Information entropy signatures in the macroscopic limit}
In this final section we explore theoretical possibilities that arise from the information entropy signatures presented in the previous section. The discussion presented here is complemented by the section \textbf{Numerical explorations on longer lattices} in the Supplementary Material.

@@

The interpretation of the information entropy signatures of the previous section in terms of information entropy wave functions opens several avenues for investigation. Here we explore its most forthright corollary, which is the establishment of uncertainty relations for topological phase transitions. The counterpart of the spatial information entropy wave function $\psi_S(x)$ in wavevector space is its Fourier transform 
\begin{equation}
\label{fourier_shannon_wave_function}
\hat{\psi}_S(k) = \int_{0}^{1}\psi_S(x)e^{-2\pi ikx}dx
\end{equation}
from which the Shannon information entropy density function can be computed in wavevector space,
\begin{equation}
\label{shannon_wave_function_wavevector_space}
\hat{\rho}_S(k) = |\hat{\psi}_S(k)|^2.
\end{equation}

Let us further denote the mean, the variance and the entropy of $\rho_S$ by $\bar{\rho}_S$, $\Delta \rho_S^2 $ and $H[\rho_S]$ respectively. Explicitly, we have 
\begin{subequations}
\label{statistics}
\begin{equation}
\label{mean}
\bar{\rho}_S = \int_0^1x\rho_S(x)dx,
\end{equation}
\begin{equation}
\label{variance}
\Delta\rho_S^2 = \int_0^1(x-\bar{\rho}_S)^2\rho_S(x)dx,\quad \text{and}
\end{equation}
\begin{equation}
\label{entropy}
H[\rho_S] = \int_0^1\rho_S(x)\ln\big(\rho_S(x)\big) dx
\end{equation}
\end{subequations}

In possession of this quantum formalism, we may write two canonical uncertainty relations binding together the real space and wavevector space information entropy density functions \eqref{shannon_wave_function} and \eqref{shannon_wave_function_wavevector_space} for topological phase transitions:

\begin{enumerate}
\item[i)] \emph{the Heisenberg uncertainty principle}
\begin{equation}
\label{heisenberg}
\Delta\rho_S \cdot \Delta\hat{\rho}_S \geq \frac{1}{4\pi}, \quad
\end{equation}
and 
\item[ii)] \emph{the Hirschman entropic uncertainty}
\begin{equation}
\label{hirschman}
H[\rho_S] + H[\hat{\rho}_S] \geq \ln\Big(\frac{e}{2}\Big). \quad
\end{equation}
\end{enumerate}

Equations \eqref{entropy_density}-\eqref{hirschman} provide a physics-grounded interpretation of the information entropy signatures discovered in the previous section from sheer data analysis of finite SSH systems in real space. In particular, the uncertainty relations \eqref{heisenberg} and \eqref{hirschman} provide rigorous measures of localizability of topological phase transitions in SSH systems based on information theory.

Given that our analysis has now come full circle back to theoretical physics, we leave the task of numerically exploring equations \eqref{entropy_density}-\eqref{hirschman} for the section \textbf{Numerical explorations on longer lattices} in the Supplementary Material.

@@

\textcolor{blue}{This paper should also contrasted with \cite{ml_interpretability}, where the subject of investigation is the interpretability of neural network models trained to recognize topological phase transitions in some physical systems. In that paper, interesting visualizations are shown demonstrating that the patterns captured by a single-layer feedforward neural network indeed correspond to known physical quantities that are relevant to the problems at hand. We agree that such tasks should indeed be called "model interpretability", as in that case the authors introspect into their models to make sure that they are learning patterns of physical pertinence to the systems being investigated. In our paper we prefer the term "model explainability", as we are using similar model introspection tools to propose previously unknown concepts and properties of the physical systems being investigated. While the nuances in the meanings of these two terms are the subject of often heated philosophical debates in the artificial intelligence community, this choice of nomenclature suits the practical application of these model introspection techniques to physics well.}\citequote{superior_comparable}

@@

\textcolor{blue}{By starting from eigenvector data generated from the simulation of SSH systems in real space, passing through proposing an approach based on eigenvector ensembling and decision trees and using model explainability to uncover the information entropy signatures presented in this article, and then exploring the theoretical possibilities posed by the information entropy signatures, our work exemplifies a full cycle of data-driven physics and illustrates how the interactions between the two disciplines can be enriching in both ways. }
